Rice yield improvements through plant breeding are offset by inherent yield declines over time

### 2018-03-11 output/breed_paper_revised.Rnw
Another big yank to conclusions.

This analysis suggests that the CCRRF breeding program has
maintained but not substantially improved the yield potential in the
face of changing biotic and abiotic pressures over the 33 year time
frame of this study. This conclusion holds
true even when evaluating the cultivars which have been breed for
broad adaption and high yield potential (M-202, M-206 and
M-105). Yields of these cultivars at time of release have all yielded
between
\Sexpr{tmp=mod_data[mod_data[,"yrs_in_trial"]<=0&mod_data[,"id"]%in%c("M202","M206","M105"),];round(min(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
    and
    \Sexpr{round(max(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
    Mg ha\tsups{-1} with no indication of yields increasing
    (Fig. \ref{fig:first_yrs}).

Yields of a particular cultivar may decline overtime for a number of
reasons. First, ~\citet{Nie2009} reported yield declines in rice due
to continuously flooded conditions over a long time period which could
be overcome with improved N management. This is likely not the case
here as the fields used in this study grow rice only one season a year
and tend to be non-flooded for the remainder of the year. Furthermore,
at this location and unlike most CA rice fields, the fields are left
fallow every other year. Second, changing climate may influence yields
with either warmer~\citep{van2013} or cooler~\citep{espe2017}
temperatures reducing yields. Again, this is unlikely to be the case
in this situation as ~\citet{espe2017} found no substantial changes in
growing season temperatures (either high, low, or average temps) the
Sacramento Valley during the study period. Thirdly, small changes in
biotic and abiotic forces have previously been speculated as the cause
for yield decline over time of the cultivar IR8 at IRRI~\citep{Peng1998,
Peng2000, Peng2010}. Since the cultivar's genetics remain stable from
the time of selection but the conditions the cultivar experiences over
time are continuously changing, accumulated changes too small to
reliably measure could plausibly lead to yield
declines over time~\citep{Peng1998, Peng2000, Peng2010, Mackay2011}. However, it is not clear
if this mechanism is responsible for the yield decline over time of
\Sexpr{yrs_coef} Mg ha\tsups{-1} year\tsups{-1} found in this
analysis. Biotic stressors such as pests and diseases are low in this
Mediterranean climate. Furthermore, these fields are fallowed every
other year and the residues are burned, further lowering the potential
for pest and disease pressure. More data is needed to further explore
the exact cause of the yield declines observed here, as well to
determine if different cultivars have differential declines over time.

It is striking that the estimated yield decline over time is similar to that
reported by \citet{Peng2010} over a similar time span (15\% decline
over roughly 30 years). This suggests the mechanisms responsible could
potentially be endemic to all rice production systems. While efforts
should be made to identify the cause of yield decline over time, this study highlights
both the need for maintenance of plant breeding and the difficulty of
increasing genetic yield ~\citep{espe2017} potential, especially in
systems such as CA where yield is approaching the physiological limit
\citep{espe16a}.

@@

@@

California state average yields have increased slowly during the study
period (Fig. \ref{fig:state_yield}) \citep{espe16a}. What are the
potential causes?  It is not due to the adoption of the broadly
adapted cultivars (M-202, M-206, and M-105) as there is no evidence of
increased yield potential among these (Fig. \ref{fig:first_yrs} \&
\ref{fig:compare_yrs}). Although M-205 is not broadly adapted, it has
increase yield potential and its adoption (although limited) in the
northern and warmer part of CA may contribute to some yield gains
reported at the state level. Other factors include improvement in
agronomic management, including more efficient nutrient
management~\citep{Lundy2012, Lundy2015, linquist2009}, and weed
management~\citep{Saito2008, Pittelkow2012, linquist2008}. However,
recent rates of state-wide yield gains are far less than those of the
preceding era, consistent with a system near the yield
ceiling~\citep{Grassini2013}.

\citet{mckenzie2014} explored the yield of rice in this same system
during a similar time period using similar data as these here, but
concluded that genetic rice yield potential has increased over this
time period at a rate of 25 kg\tsups{-1} ha\tsups{-1}
year\tsups{-1}. Yet the CI estimated from these data using the best
model suggests that there is only a
\Sexpr{round(sum((raw_samples$release_yr_c*scl)>=25)/nrow(raw_samples)*100,1)}\%
probability of the rate of yield increase is equal to or greater than
25 kg\tsups{-1} ha\tsups{-1} year\tsups{-1}. There are several
explanations for this discrepancy in conclusions. First, \citet{mckenzie2014} did not
account for yield decreases over time in their analysis. As shown
here, when these decreases are not taken into about, the estimated
increase in yield over time is estimated to be approximately the same
as that found by \citet{mckenzie2014}. Second, these yield decreases over time
were not taken into account possibly because \citet{mckenzie2014}
concluded that there was no trend in the yield of the major cultivars
over time. However this conclusion is based on a flawed interpretation
of the results of their statistical test. Failing to reject the
hypothesis that the trend in yield equals zero does not equate to the
confirmation of no trend. Instead, failing to reject the hypothesis
could suggest the analysis lacks the statistical power to detect the
trend. Third, it
is not surprising that no trend in yield over time was detected given
that \citet{mckenzie2014} used data from several locations, yet
analyzed those data using a simple t-test. By using this simple
analysis, \citet{mckenzie2014} ignore systematic errors related to
differences between observations from different locations, including
different management systems, growing environments, and crop
timing. By contrast, the investigation presented here used a single,
well managed system and analyzed the data using a mixed effects model
which fully accounts for non-independence between similar
observations. In doing this, the analysis makes fewer assumptions and
more fully captures the structural relationships in the data. Hence,
the estimates presented here are likely more reliable.

@@

>>>>

This analysis suggests that the CCRRF breeding program has
maintained but not substantially improved the yield potential in the
face of changing biotic and abiotic pressures over the 33 year time
frame of this study. This conclusion holds
true even when evaluating the cultivars which have been breed for
broad adaption and high yield potential (M-202, M-206 and
M-105).

\subsection{Causes of declining yields}

Yields of a particular cultivar may decline overtime for a number of
reasons. First, ~\citet{Nie2009} reported yield declines in rice due
to continuously flooded conditions over a long time period which could
be overcome with improved N management. This is likely not the case
here as the fields used in this study grow rice only one season a year
and tend to be non-flooded for the remainder of the year. Furthermore,
at this location and unlike most CA rice fields, the fields are left
fallow every other year. Secondly, small changes in biotic and abiotic
forces have previously been speculated as the cause for yield decline
over time of the cultivar IR8 at IRRI~\citep{Peng1998, Peng2000,
  Peng2010}. Since the cultivar's genetics remain stable from the time
of selection but the conditions the cultivar experiences over time are
continuously changing, accumulated changes too small to reliably
measure could plausibly lead to yield declines over
time~\citep{Peng1998, Peng2000, Peng2010, Mackay2011}. However, it is
not clear if this mechanism is responsible for the yield decline over
time of \Sexpr{yrs_coef} Mg ha\tsups{-1} year\tsups{-1} found in this
analysis. Biotic stressors such as pests and diseases are low in this
Mediterranean climate. Furthermore, these fields are fallowed every
other year and the residues are burned, further lowering the potential
for pest and disease pressure. Thirdly, changing climate may influence
yield potential with either warmer~\citep{van2013} or
cooler~\citep{espe2017} temperatures reducing
yields. Yield potential can be defined as the yield of optimum
cultivar in the absence of biotic and abiotic stresses in a given
environment \citep{vanIttersum2013}. However, ~\citet{espe2017} found
no substantial trends in growing season temperatures (either high,
low, or average temps) in the Sacramento Valley during the study
period, suggesting that older cultivars do not have lower yield
potential due to changing climatic factors. While our analysis
indicates that there is no interaction between cultivar and year (CI
includes zero), visual examination suggests that cultivars may be
different in terms of rates of yield decline
(Fig. \ref{fig:all_cult}); however, we do not have sufficient data nor
the proper controls in the existing data set to fully test this
hypothesis. More data is needed to further explore the exact cause of
the yield declines observed here, as well to determine if different
cultivars have differential declines over time.

\subsection{Causes for statewide increases over time}

\begin{comment}
It is striking that the estimated yield decline over time is similar
to that reported by \citet{Peng2010} over a similar time span (15\%
decline over roughly 30 years). This suggests the mechanisms
responsible could potentially be endemic to all rice production
systems. While efforts should be made to identify the cause of yield
decline over time, this study highlights both the need for maintenance
of plant breeding and the difficulty of increasing genetic yield
~\citep{espe2017} potential, especially in systems such as CA where
yield is approaching the physiological limit \citep{espe16a}.

@@

California state average yields have increased slowly during the study
period (Fig. \ref{fig:state_yield}) \citep{espe16a}. What are the
potential causes?  It is not due to the adoption of the broadly
adapted cultivars (M-206 and M-105) as there is no evidence of
increased yield potential among these (Fig. \ref{fig:first_yrs} \&
\ref{fig:compare_yrs}) compared to M-202 released at the start of this
period. Although M-205 is not broadly adapted, it has
increase yield potential (Fig. \ref{fig:first_yrs} and its adoption (although limited) in the
northern and warmer part of CA may contribute to some yield gains
reported at the state level. Other factors include improvement in
agronomic management, including more efficient nutrient
management~\citep{Lundy2012, Lundy2015, linquist2009}, and weed
management~\citep{deforest2017, Pittelkow2012, caton2002}. However,
recent rates of state-wide yield gains are far less than those of the
preceding era, consistent with a system near the yield
ceiling~\citep{Grassini2013}.

\begin{comment}
\citet{mckenzie2014} explored the yield of rice in this same system
during a similar time period using similar data as these here, but
concluded that genetic rice yield potential has increased over this
time period at a rate of 25 kg\tsups{-1} ha\tsups{-1}
year\tsups{-1}. Yet the CI estimated from these data using the best
model suggests that there is only a
\Sexpr{round(sum((raw_samples$release_yr_c*scl)>=25)/nrow(raw_samples)*100,1)}\%
probability of the rate of yield increase is equal to or greater than
25 kg\tsups{-1} ha\tsups{-1} year\tsups{-1}. There are several
explanations for this discrepancy in conclusions. First, \citet{mckenzie2014} did not
account for yield decreases over time in their analysis. As shown
here, when these decreases are not taken into about, the estimated
increase in yield over time is estimated to be approximately the same
as that found by \citet{mckenzie2014}. Second, these yield decreases over time
were not taken into account possibly because \citet{mckenzie2014}
concluded that there was no trend in the yield of the major cultivars
over time. However this conclusion is based on a flawed interpretation
of the results of their statistical test. Failing to reject the
hypothesis that the trend in yield equals zero does not equate to the
confirmation of no trend. Instead, failing to reject the hypothesis
could suggest the analysis lacks the statistical power to detect the
trend. Third, it
is not surprising that no trend in yield over time was detected given
that \citet{mckenzie2014} used data from several locations, yet
analyzed those data using a simple t-test. By using this simple
analysis, \citet{mckenzie2014} ignore systematic errors related to
differences between observations from different locations, including
different management systems, growing environments, and crop
timing. By contrast, the investigation presented here used a single,
well managed system and analyzed the data using a mixed effects model
which fully accounts for non-independence between similar
observations. In doing this, the analysis makes fewer assumptions and
more fully captures the structural relationships in the data. Hence,
the estimates presented here are likely more reliable.
\end{comment}

\subsection{Other advances from breeding}

@@

@@

Increasing grain yield is essential to meeting the challenges of
feeding a growing population in a changing world. Here we show that
side-by-side cultivar comparisons do not necessarily indicate gains in
yield potential as they do not account for yield declines over time of
the older cultivars being tested. Here we show, as have others, that
the yield of a cultivar declines over time. Therefore, quantifying
changes in yield potential requires comparing yields of new cultivar
yields to yields of the older cultivars when they were first
released. Unfortunately, our analysis did not allow us to identify the
cause of yield declines in this case. While efforts should be made to
identify the cause of yield decline over time, this study highlights
both the need for plant breeding to simply maintain current yields, as
well as the challenge of increasing genetic yield potential,
especially in systems such as CA where yield is approaching the
physiological limit. The goal of increasing yield may require the
yield performance of cultivars to be assessed earlier in the selection
process to avoid discarding higher yielding entries without full
consideration. We suggest here that historical benchmarks be used in
to evaluate promising rice cultivars in addition to side-by-side
comparisons to maintain perspective of the absolute yield gains being
made.
